---
title: "Modeling Plant Establishment from Ornithochory"
author: "David S, Mason"
date: "4/15/2021"
output:
  pdf_document: default
  html_document: default
---

## Overview

*Animals track resources*

*Animals disperse plants*

*Disturbance creates resource pulses*

*Disturbance creates favorable conditions for plant establishment*

**HO: The number of plants establishing via ornithochory is the same in recent burns and 1 year rough.**

**HA: The number of plants establishing via ornithochory is different in recent burns than 1 year rough.**

Below is a series of models using seed rain and plant establishment data in conjunction to predict the number of plants establishing from bird-mediated rain at recent burns and 1 year rough. *This model does not include density-dependent effects, which are known to influence plant population dynamics.*

First, I predict mean seed arrival per m^2 (below a perch). Next, I simulate plant establishment trials (these are not completed). Then, I multiple the number of arrivng seeds by the probability of establishment to generare an estimated PDF of establshing individuals. Finally, I develop a framework for exploring the effects of variation in lambda and p.


## Set-up the workspace
```{r, warning=FALSE,message=FALSE}
library(tidyverse) # load in Tidyverse to filter data

```
```{r setup}
require(tidyverse)

# Load in the data to filter data
seeds <- read.csv("seed_trap.csv", header = T) # Bring in the data

# Convert the site x species matrix into the long format
seeds_long <- pivot_longer(seeds, cols = 4:9,names_to = "SPECIES", 
													 values_to = "SEEDS")

# Filter the data by treatment (control is 1 year rough)
burn <- seeds_long %>% 
filter(TREATMENT == "BURN")
control <- seeds_long %>% 
filter(TREATMENT == "CONTROL")
```

## Calculate seed arrival MLE
```{r}
treatment <- function(burn,control ,plot.it=TRUE){

	# Computing Likelihood for Observed Data ####
	llh_poisson <- function(lambda, y){
		# log(likelihood) by summing 
		llh <- sum(dpois(y, lambda, log=TRUE))
  		return(llh)
	}

	lambdas <- seq(0,100, by=0.01)

	# compute log-likelihood for all lambda values
	burn_ll <- sapply(lambdas,function(x){llh_poisson(x,burn$SEEDS)})
	max.burn.ll <- max(burn_ll)
	burn.mle <- lambdas[burn_ll==max.burn.ll]
	burn.rel.l <- exp(burn_ll - max.burn.ll)
	burn.ci.cuttoff <- exp(-qchisq(p=0.95, df=1)/2)
	burn.cut.m.rell <- burn.ci.cuttoff - burn.rel.l
	burn.ci.indices <- which(burn.cut.m.rell<0, arr.ind=TRUE)
	burn.len.ciind <- length(burn.ci.indices)
	burn.left.CL <- lambdas[burn.ci.indices[1]]
	burn.right.CL <-lambdas[burn.ci.indices[burn.len.ciind]]
	n.burn <- length(burn$SEEDS)
	
	control_ll <- sapply(lambdas,function(x){llh_poisson(x,control$SEEDS)})
	max.control.ll <- max(control_ll)
	control.mle <- lambdas[control_ll==max.control.ll]
	control.rel.l <- exp(control_ll - max.control.ll)
	control.ci.cuttoff <- exp(-qchisq(p=0.95, df=1)/2)
	control.cut.m.rell <- control.ci.cuttoff - control.rel.l
	control.ci.indices <- which(control.cut.m.rell<0, arr.ind=TRUE)
	control.len.ciind <- length(control.ci.indices)
	control.left.CL <- lambdas[control.ci.indices[1]]
	control.right.CL <-lambdas[control.ci.indices[control.len.ciind]]
	n.control <- length(control$SEEDS)	

	#### Null model log-likelihood: no difference between burn and control?
	all.data  <- c(as.vector(burn$SEEDS), as.vector(control$SEEDS))
	null_ll <- sapply(lambdas,function(x){llh_poisson(x,all.data)})
	max.null.ll <- max(null_ll)
	null.mle <- lambdas[null_ll==max.null.ll]

	bic.sep.model <- -2*(max.burn.ll+max.control.ll) + 2*log(n.burn+n.control)
	bic.null.model <- -2*max.null.ll + 1*log(n.burn+n.control)
	if(bic.sep.model<bic.null.model){outcome <- 
		"Separate model is best"}else{outcome<- 
			"There's no difference between control and burn"}
	# Save the lambdas and log-likelihoods in a data frame
	burn_df <- data.frame(ll=burn_ll, lambda=lambdas)
	burn_df$TREATMENT <- c("BURN")
	control_df <- data.frame(ll=control_ll, lambda=lambdas)
	control_df$TREATMENT <- c("CONTROL")
	profile.ll <- rbind(burn_df, control_df)
	
	# Save the MLES and the confidence intervals	
	burn.cis <- c(burn.left.CL, burn.mle, burn.right.CL)
	ctl.cis  <- c(control.left.CL, control.mle, control.right.CL)
	out <- rbind(burn.cis, ctl.cis)	
	colnames(out) <- c("LCL", "MLE", "UPL")
	rownames(out) <- c("Burn treat.", "Control")
	
	
	
	if(plot.it==TRUE){
		
		burn.ci <- seq(from=burn.left.CL, to=burn.right.CL, by=0.01)
		ctl.ci <- seq(from=control.left.CL, to=control.right.CL, by=0.01)
		ci.height <- burn.ci.cuttoff
		
		xlims.burn <- c(burn.left.CL-1, burn.right.CL+1)
		xlims.ctl  <- c(control.left.CL-1, control.right.CL+1) 
		par(mfrow=c(1,2), oma=c(1,2,1,1), mar=c(4,5,4,2))		
		plot(lambdas, burn.rel.l, type="l", col="red", lwd=2, xlim=xlims.burn, xlab=expression(lambda), 
		ylab="Relative Profile Likelihood", cex.lab=1.5, bty="l", main="Burn treatment")
		polygon(x=c(burn.ci,rev(burn.ci)), y=c(rep(0,length(burn.ci)),rep(ci.height,length(burn.ci))), col="grey")
		plot(lambdas, control.rel.l, type="l", col="red", lwd=2, xlim=xlims.ctl,xlab=expression(lambda), 
		ylab="Relative Profile Likelihood", cex.lab=1.5, bty="l", main="Control")
		polygon(x=c(ctl.ci,rev(ctl.ci)), y=c(rep(0,length(ctl.ci)),rep(ci.height,length(ctl.ci))), col="grey")		
		
	}
	
	
	return(list(CIS.lambdas=out,bic.sep.model=bic.sep.model, bic.null.model=bic.null.model, outcome=outcome))
}
```
```{r, echo=FALSE}
output <- treatment(burn,control,plot.it=TRUE)
```
```{r}
# Extract the MLEs
out.mle <- output[[1]]

# Separate the burn from control MLEs
burn.lamb.mle <- out.mle[1,3]
control.lamb.mle <- out.mle[2,3]
```

## Simulate plant establishment data
```{r}
# Number of trials 
N <- 10
burn.reps <- 21
control.reps <- 21

# Simulate the data
burn.samp <- rbinom(n=burn.reps, size=N, prob=0.40)
cont.samp <- rbinom(n=burn.reps, size=N, prob=0.16)

# Calculate phat
burn.p.hat <- sum(burn.samp)/(N*length(burn.samp))
cont.p.hat <- sum(cont.samp)/(N*length(cont.samp))
```

## Generate joint PDF using mean seed arrival and probability of establishment 
```{r}
# Create a vector
x <- seq(1,20)

# Generate PDF 
burn.est <- dpois(lambda=floor((burn.lamb.mle*burn.p.hat)), x=x)
con.est <- dpois(lambda=floor((control.lamb.mle*cont.p.hat)), x=x)

# Convert into data frame
burn.est <- as.data.frame(cbind(x,burn.est))
con.est <- as.data.frame(cbind(x,con.est))

# Add treatment and MLE column
burn.est$Treatment <- "Burn"
con.est$Treatment <- "Control"

burn.est$MLE <- burn.lamb.mle*burn.p.hat
con.est$MLE <- control.lamb.mle*cont.p.hat

burn.est <- setNames(burn.est,c("Plants","Probability",
																"Treatment", "MLE"))
con.est <- setNames(con.est,c("Plants","Probability",
															"Treatment", "MLE"))

# Combine the dataframes
est.dat <- rbind(burn.est,con.est)
head(est.dat)
```
```{r, echo=FALSE}
ggplot(d=est.dat, aes(x=Plants, y=Probability, color = Treatment))+
	geom_line(size=1.5)+
	xlim(0,15)+
	scale_color_manual(values=c("#22A884FF","#FCA636FF"))+
	theme_bw()+
	geom_vline(data=est.dat, aes(xintercept = MLE))+
	facet_wrap(~Treatment)
```

## Exploring different values for lambda / sample sizes
```{r}
change.inc <- c(0.25, 0.50, 0.75)
change.dec <- c(-0.75,-0.50,-0.25)

burn.samp.inc <- c()
burn.samp.dec <- c()

for(i in 1:3){
	burn.samp.inc[i] <- abs((burn.lamb.mle*change.inc[i])+burn.lamb.mle)
	burn.samp.dec[i] <- abs((burn.lamb.mle*change.dec[i])+burn.lamb.mle)
}

burn.samp <- c(burn.samp.dec,burn.lamb.mle,burn.samp.inc)

## Percent changes from the seed arrival MLE for control

con.samp.inc <- c()
con.samp.dec <- c()

for(i in 1:3){
	con.samp.inc[i] <- abs((control.lamb.mle*change.inc[i])+control.lamb.mle)
	con.samp.dec[i] <- abs((control.lamb.mle*change.dec[i])+control.lamb.mle)
}

con.samp <- c(con.samp.dec,control.lamb.mle,con.samp.inc)

### For loop to generate PDFs with different lambda values

pdf.list <-list()
x <- seq(1,50)

for(i in 1:length(burn.samp)){
	burn.pdf <- dpois(lambda=floor((burn.samp[i]*burn.p.hat)), x=x)
	con.pdf <- dpois(lambda=floor((con.samp[i]*cont.p.hat)), x=x)
	
	# Convert into data frame
	burn.pdf <- as.data.frame(cbind(x,burn.pdf))
	con.pdf <- as.data.frame(cbind(x,con.pdf))

	# Add treatment and MLE column
	burn.pdf$Treatment <- "Burn"
	con.pdf$Treatment <- "Control"

	burn.pdf$MLE <- burn.samp[i]*burn.p.hat
	con.pdf$MLE <- con.samp[i]*cont.p.hat

	burn.pdf <- setNames(burn.pdf,c("Plants","Probability",
																"Treatment", "MLE"))
	con.pdf <- setNames(con.pdf,c("Plants","Probability",
																"Treatment", "MLE"))
# Combine the dataframes
	pdf <- rbind(burn.pdf,con.pdf)
	pdf.list[[i]] <- pdf
}

#### Pull
pdf1 <- pdf.list[[1]]
pdf1$Change <- "-75%"

pdf2 <- pdf.list[[2]]
pdf2$Change <- "-50%"

pdf3 <- pdf.list[[3]]
pdf3$Change <- "-25%"

pdf4 <- pdf.list[[4]]
pdf4$Change <- "No change"

pdf5 <- pdf.list[[5]]
pdf5$Change <- "+25%"

pdf6 <- pdf.list[[6]]
pdf6$Change <- "+50%"

pdf7 <- pdf.list[[7]]
pdf7$Change <- "+75%"

all.pdf <- rbind(pdf1,pdf2,pdf3,pdf5,pdf6,pdf7)
all.pdf$Change <- as_factor(all.pdf$Change)
all.pdf$Change <- factor(all.pdf$Change, levels = c("-75%", "-50%","-25%",
																		 							"+25%","+50%","+75%"))
```
```{r, warning=FALSE,message=FALSE}
ggplot(d=all.pdf, aes(x=Plants, y=Probability, color = Treatment))+
	geom_line(size=1)+
	scale_color_manual(values=c("#22A884FF","#FCA636FF"))+
	theme_bw()+
	facet_wrap(~Change+Treatment)
```